\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

As I pen down the final page of this thesis, I am overwhelmed with gratitude for the transformative past three years. It has been a journey filled with challenges, doubts, and substantial learning in between. The growth I have experienced and the completion of this thesis would not have been possible without the continuous support, guidance, and encouragement from many.

First and foremost, my deepest gratitude goes to my advisors, Guillaume Wisniewski and Benoît Crabbé, to whom I owe nearly everything I know about research. I really appreciate the enlightening discussions during our weekly meetings, reading groups, and even over lunch breaks. Benoît, I profoundly admire your expertise in both linguistics and NLP. Thank you for your insightful and detailed explanations of scientific concepts, and for helping me to find my own research path. Guillaume, you introduced me to the academic world, guiding me step by step in research. I am grateful for your trust and sense of humor, which have carried me through many challenging moments. You ingrained in me the belief that everything is learnable, a lesson that has deeply shaped me into a better version of myself. 

I am grateful to Tal Linzen and Najoung Kim for their invaluable mentorship during my visit to NYU. My appreciation extends to the NYU Computation and Psycholinguistics Lab team for very helpful discussion and to Alexander, Yuekun, and Lucia for their collaboration. I also thank Labex EFL and LLF for funding this opportunity.

Dear committee members, I appreciate the time and effort you have invested in reviewing my thesis and accompanying me during the final stage of my PhD.

To the friends and colleagues from my PhD journey, your help and support have been crucial to me. Special thanks to Nathanaël, Juliette, Antoine, Anna, David, Julie, Marie, Timothée, Huiyi, Clémentine, Yiming, Fang, Aixiu, Chuyuan, Maria, Zhanglin, Dorotea, Ensieh, Saida, Zulipiye. I extend my heartfelt gratitude to all the members of LLF for their invaluable support.

A warm thank you to my husband, Bo, a programmer, for your support, and patience through my numerous practice talks. 

Lastly, I am grateful to my parents for instilling in me the courage and wisdom to pursue my passions. I still remember a moment from my childhood during the busy harvest season; as I stood before the seemingly endless rice fields awaiting harvest, I felt overwhelmed and scared. You looked at the sickle in my hands and reassured me, ``Just start with the stalk of rice at your feet, and we will get through the entire field''. This simple phrase has stayed with me as I ventured beyond our mountain village, beyond our small county, into university, and out into the vast world, surmounting numerous challenges. I aspire that my future work will, in some way, connect back to the land that my parents cherish so dearly. 